---
title: "Chapter 3"
output: html_document
---
```{r}
library(rethinking)
```
# Easy

Model definition
$$
y_{i} \sim Normal(\mu,\sigma)
$$
$$
\mu \sim Normal(0, 10)
$$
$$
\sigma \sim Exponential(1)
$$

1. Line 1 is the likelihood

2. 2 parameters in the posterior distribution, `\sigma` and `\mu`

3. Write down the appropriate form of Bayes' theorem that includes the proper likelihood and priors

$$
Pr(\mu,\sigma | y) = \frac{\prod_{i}{Normal(y_{i}|\mu,\sigma)Normal(\mu|0,10)Exponential(\sigma|1)}}{\int{\int{\prod_{i}{Normal(y_{i}|\mu,\sigma)Normal(\mu|0,10)Exponential(\sigma|1)}}d\mu d\sigma}}
$$

4. 

$$
y_{i} \sim Normal(\mu,\sigma)
$$
$$
\mu_{i} = \alpha + \beta x_{i}
$$
$$
\alpha \sim Normal(0, 10)
$$
$$
\beta \sim Normal(0,1)
$$
$$
\sigma \sim Exponential(2)
$$
Line 2 is the linear model.

5. There are 3 + `i` parameters? $$\alpha$$, $$\beta$$, $$\sigma$$ and each $$\mu_{i}$$? Or is $$\mu$$ count as one?

## Medium

1. Simulate observed `y` values from the prior (not the posterior) for the definition

$$
y_{i} \sim Normal(\mu,\sigma)
$$
$$
\mu \sim Normal(0, 10)
$$
$$
\sigma \sim Exponential(1)
$$

```{r}
curve(dnorm(x, 0, 10), from = -40, to = 40)
```

```{r}
curve(dexp(x, 1), from = 0, to = 10)
```

```{r}
sample_mu <- rnorm(1e4, 0, 10)
sample_sigma <- rexp(1e4, 1)
prior_y <- rnorm(1e4, sample_mu, sample_sigma)
dens(prior_y)
```

2. Translate the model into a `quap` formula

```{r}
flist <- alist(
  y ~ dnorm(mu, sigma),
  mu ~ dnorm(0, 10),
  sigma ~ dexp(1)
)
```

3. Translate the `quap` formula into a mathematical model definition

```{r}
ex <- alist(
  y ~ dnorm( mu , sigma ),
  mu <- a + b*x,
  a ~ dnorm( 0 , 10 ),
  b ~ dunif( 0,1),
  sigma ~ dexp( 1 )
)
```

$$
y_{i} \sim Normal(\mu, \sigma)
$$
$$
\mu = \alpha + \beta x
$$
$$
\alpha \sim Normal(0, 10)
$$
$$
\beta \sim Uniform(0, 1)
$$
$$
\sigma \sim Exponential(1)
$$

4. A sample of students is measured for height each year for 3 years. After the 3rd year, you want to fit a linear regression predicting height using year as a predictor. Write down the mathematical model definition for this regression, using any variable names and priors you choose. Be prepared to defend your choice of priors.

$$
h_{i} \sim Normal(\mu, \sigma)
$$
$$
\mu = \alpha + \beta y
$$
$$
\alpha \sim Normal(178, 20)
$$
$$
\beta \sim Exponential(0.25)
$$
$$
\sigma \sim Uniform(0, 50)
$$
Going for exponential beta as I expect that height in students will increase year on year, so want to ignore <0. I think ideal world could have a prior that wouldn't have it's mode at 0 as they are more likely to grow but thought this would be close enough. Use rate 0.25 as made the curve look "not too flat"
```{r}
curve(dexp(x, rate = 0.25), from = 0, to = 10)
```
5. Doesn't change as I tried to factor that into part 4

6. We have alpha is normally disributed with $$\sigma^{2} = 20$$, if within a single age never vary by more than 64cm, we can use this to change $$\alpha$$. In normal, 95% of the data is within 2sds of the mean. So we could put 95% bounds at 64cm. So `(a + 2sd) - (a - 2sd) = 64` => `sd = 16`

```{r}
curve(dnorm(x, mean = 178, sd = 16), from = 130, to = 230)
```
Or would we revise to 64cm to cover 99.7% by using 3 sds from the mean? Would give tighter prior

```{r}
curve(dnorm(x, mean = 178, sd = 16), from = 130, to = 230)
```

7. Refit model m4.3, but omit the mean weight `xbar`. Compare the new posterior to the original model, look at covariance among parameters. Compare posterior predictions

```{r}
library(rethinking)

data(Howell1)

d <- Howell1
d2 <- d[d$age >= 18, ]

```


```{r}
plot(NULL, 
     xlim=range(d2$weight), 
     ylim=c(-100, 400),
     xlab="weight", 
     ylab="height")

abline( h=0 , lty=2 )

abline( h=272 , lty=1 , lwd=0.5 )

mtext( "b ~ dnorm(0,10)" )

xbar <- mean(d2$weight)

N <- 100
a <- rnorm(N, 178, 20)
b <- runif(N, 0, 1)
print(b)
for (i in 1:N) {
  curve(a[i] + b[i]*(x),
        from = min(d2$weight), 
        to = max(d2$weight), 
        add = TRUE,
        col = col.alpha("black", 0.2))
}
```
```{r}
# fit <- quap(
#   alist(
#     height ~ dnorm(mu, sigma),
#     mu <- a + b * weight,
#     a ~ dnorm(178, 20),
#     b ~ dunif(0, 1),
#     sigma ~ dunif(0, 5)
#   ),
#   data = d2
# )
```
