---
title: "Chapter 3"
output: html_document
---
```{r}
library(rethinking)
```
# Easy

Model definition
$$
y_{i} \sim Normal(\mu,\sigma)
$$
$$
\mu \sim Normal(0, 10)
$$
$$
\sigma \sim Exponential(1)
$$

1. Line 1 is the likelihood

2. 2 parameters in the posterior distribution, `\sigma` and `\mu`

3. Write down the appropriate form of Bayes' theorem that includes the proper likelihood and priors

$$
Pr(\mu,\sigma | y) = \frac{\prod{Normal(y_{i}|\mu,\sigma)Normal(\mu|0,10)Exponential(\sigma|1)}}{\int{\int{Normal(y_{i}|\mu,\sigma)Normal(\mu|0,10)Exponential(\sigma|1)}}d\mu d\sigma}
$$

4. 

$$
y_{i} \sim Normal(\mu,\sigma)
$$
$$
\mu_{i} = \alpha + \beta x_{i}
$$
$$
\alpha \sim Normal(0, 10)
$$
$$
\beta \sim Normal(0,1)
$$
$$
\sigma \sim Exponential(2)
$$
Line 2 is the linear model.

5. There are 3 + `i` parameters? $$\alpha$$, $$\beta$$, $$\sigma$$ and each $$\mu_{i}$$? Or is $$\mu$$ count as one?

## Medium

1. Simulate observed `y` values from the prior (not the posterior) for the definition

$$
y_{i} \sim Normal(\mu,\sigma)
$$
$$
\mu \sim Normal(0, 10)
$$
$$
\sigma \sim Exponential(1)
$$

```{r}
curve(dnorm(x, 0, 10), from = -40, to = 40)
```

```{r}
curve(dexp(x, 1), from = 0, to = 10)
```

```{r}
sample_mu <- rnorm(1e4, 0, 10)
sample_sigma <- rexp(1e4, 1)
prior_y <- rnorm(1e4, sample_mu, sample_sigma)
dens(prior_y)
```

2. Translate the model into a `quap` formula

```{r}
flist <- alist(
  y ~ dnorm(mu, sigma)
  mu ~ dnorm(0, 10)
  sigma ~ dexp(1)
)
```

3. Translate the `quap` formula into a mathematical model definition

```{r}
ex <- alist(
  y ~ dnorm( mu , sigma ),
  mu <- a + b*x,
  a ~ dnorm( 0 , 10 ),
  b ~ dunif( 0,1),
  sigma ~ dexp( 1 )
)
```

$$
y_{i} \sim Normal(\mu, \sigma)
$$
$$
\mu = \alpha + \beta x
$$
$$
\alpha \sim Normal(0, 10)
$$
$$
\beta \sim Uniform(0, 1)
$$
$$
\sigma \sim Exponential(1)
$$

4. A sample of students is measured for height each year for 3 years. After the 3rd year, you want to fit a linear regression predicting height using year as a predictor. Write down the mathematical model definition for this regression, using any variable names and priors you choose. Be prepared to defend your choice of priors.

$$
h_{i} \sim Normal(\mu, \sigma)
$$
$$
\mu = \alpha + \beta y
$$
$$
\alpha \sim 
$$

